A Heuristic Approach towards Drawings of Graphs with High Crossing Resolution

### 2018-06-04 paper/paper.tex
Working a bit on the description

In this section, we describe our heuristic algorithm for obtaining drawings with high crossing resolution. We assume that the input of our algorithm consists of a graph $G$ and an initial drawing $\Gamma_0$ of $G$ that is of some crossing resolution $c(\Gamma_0)$. We also assume that no two edges of $G$ overlap in $\Gamma_0$, that is, $c(\Gamma_0)>0$. A circular drawing of $G$ or a drawing obtained by applying a standard force-directed algorithm on $G$ clearly meets these preconditions.

Our algorithm is iterative and at each iteration step performs some operations that are mainly based on randomization. More precisely, at the $i$-th iteration step, we assume that we have computed a drawing $\Gamma_{i-1}$ of some crossing resolution $c(\Gamma_{i-1}) \geq c(\Gamma_0)$, where $\Gamma_0$ is our initial drawing. In other words, we assume, as an invariant property of our algorithm, that the crossing resolution cannot be decreased at some iteration step. Then, a vertex of $\Gamma_{i-1}$ is chosen arbitrarily at random based on the so-called \emph{vertex-pool}, which may contain:

@@

Intuitively, the critical vertices are the endpoints of the edges that define the crossing resolution of drawing $\Gamma_{i-1}$. To formally define them, we first need to introduce the notion of critical edge-pairs. A pair of edges, say $e$ and $e'$, is called \emph{critical} in drawing $\Gamma_{i-1}$, if $e$ and $e'$ cross in $\Gamma_{i-1}$ and the minimum angle that is formed at their crossing point is equal to the crossing resolution $c(\Gamma_{i-1})$ of drawing $\Gamma_{i-1}$. The set of critical vertices of drawing $\Gamma_{i-1}$ is then defined by the four endvertices of each critical edge-pair.

The role of critical vertices is central in our algorithm\footnote{If the focus is not on the critical vertices for a large graph, then an algorithm that is based on randomly selecting a vertex to move, so to improve the crossing resolution, will need a large number of iterations in order to converge to a good solution, because it is simply very unlike to select one of them to move.}. The reason is that by slightly changing the location of a critical vertex or the location of one of its neighbours appropriately, one naturally expects to improve the crossing resolution of the current drawing. We turned this observation into an algorithmic implementation through a weighted random selection procedure, in which the vertices which are at distance $i$ from the vertices of the vertex-pool in the graph have higher weights than the corresponding vertices at distance $j$, when $ \leq <i<j$. Hence, if the vertex-pool contains critical vertices, then the closer a vertex is to a critical vertex, the more likely it is to be chosen. Otherwise, each vertex of the graph can be chosen with the same probability.

With this procedure, what we quickly realized from our experimental evaluation, is that by appropriately changing the location of a critical vertex at each iteration of our algorithm, the crossing resolution of the initial drawing improves rapidly during the first iterations of the algorithm. However, by focusing only at the critical vertices of the drawing, it is highly possible that the algorithm will get trapped to some local maxima after a number of iterations. Hence, special care is needed in order to avoid these bottlenecks, especially when the graph to be drawn is large. We discuss heuristics to avoid such bottlenecks in  Section~\ref{ssec:maxima}.

>>>>

In this section, we describe our heuristic for obtaining drawings with high crossing resolution. The input of our heuristic consists of a graph $G$ and an initial drawing $\Gamma_0$ of $G$ with crossing resolution $c(\Gamma_0)$. We assume that no two edges of~$G$ overlap in $\Gamma_0$, i.e., $c(\Gamma_0)>0$. A circular drawing or a drawing obtained by applying a force-directed algorithm on $G$ clearly meets this precondition.

Our algorithm is iterative and at each iteration performs some operations that are mainly based on randomization. At the $i$-th iteration, we assume that we have computed a drawing $\Gamma_{i-1}$ of crossing resolution $c(\Gamma_{i-1}) \geq c(\Gamma_0)$. 
%, where $\Gamma_0$ is our initial drawing. 
In other words, we assume, as an invariant for our algorithm, that the crossing resolution cannot be decreased at some iteration. Then, a vertex of $\Gamma_{i-1}$ is chosen arbitrarily at random based on the so-called \emph{vertex-pool}, which may contain:

@@

Intuitively, the critical vertices are the endpoints of the edges that define the crossing resolution of drawing $\Gamma_{i-1}$. To formally define them, we first need to introduce the notion of critical edge-pairs. A pair of edges $e$ and $e'$ is called \emph{critical} in $\Gamma_{i-1}$, if $e$ and $e'$ cross in $\Gamma_{i-1}$ and the minimum angle that is formed at their crossing point is equal to $c(\Gamma_{i-1})$. The set of critical vertices of $\Gamma_{i-1}$ is then defined by the four endvertices of each critical edge-pair.

The role of critical vertices is central in our algorithm\footnote{If the focus is not on the critical vertices for a large graph, then an algorithm that is based on randomly selecting a vertex to move, so to improve the crossing resolution, will need a large number of iterations in order to converge to a good solution, because it is simply very unlike to select one of them to move.}: By appropriately changing the location of a critical vertex or of a vertex in the neighbourhood of the critical vertices, we naturally expect to improve the crossing resolution of the current drawing. We turned this observation into an algorithmic implementation through a weighted random selection procedure, in which the vertices at distance~$i$ from the ones of the vertex-pool in the graph have higher weights than the corresponding ones at distance $j$, when $0 \leq i<j$. So, if the vertex-pool contains critical vertices, then the closer a vertex is to the critical vertices, the more likely it is to be chosen. Otherwise, each vertex of the graph can be chosen with the same probability.

%With this procedure, what we quickly realized from our experimental evaluation, is that by appropriately changing the location of a critical vertex at each iteration of our algorithm, the crossing resolution of the initial drawing improves rapidly during the first iterations of the algorithm. 
What we quickly realized from our practical analysis, is that the crossing resolution of the initial drawing improves rapidly during the first iterations of the algorithm. However, by focusing exclusively at the critical vertices, it is~highly possible that the algorithm will get trapped to some local maxima after a number of iterations. So, special care is needed to avoid these bottlenecks, especially when the graph to be drawn is large. We discuss ways to avoid them in  Section~\ref{ssec:maxima}.
